---
title: "Analyzing L.A. County's changing languages in Census microdata"
output: github_document
---

**By [Ryan Menezes](https://twitter.com/ryanvmenezes/)**

This file goes over the data and R code used to analyze the changes in L.A. County's spoken languages for the Feb. 17, 2020 L.A. Times story on [local libraries' attempts to supply books in languages other than English](https://www.latimes.com/la-me-libraries-languages-diversity-2019-story.html) to an ever-changing population.

It uses the following R packages:

```{r message=FALSE}
library(ipumsr)
library(tidyverse)
```

This analysis uses a custom microdata extract from IPUMS-USA. If you need an introduction to microdata, I recommend this [excellent write-up](https://source.opennews.org/articles/when-and-how-use-census-microdata/) by Robert Gebeloff.

The data extract comes with two files:

1. A fixed-width data file in a compressed folder
2. A data codebook in an XML format, which describes the data based on the [Data Documentation Initative](https://ddialliance.org/), or "DDI"

The `ipumsr` package can read in both, provided they have equivalent names.

```{r}
ddi = read_ipums_ddi('data/usa_00011.xml')
data = read_ipums_micro(ddi)
```

## The codebook

The DDI details the variables in the extract:

```{r}
info.variables = ddi$var_info
info.variables
```

The key column in this extract is [LANGUAGE](https://usa.ipums.org/usa-action/variables/LANGUAGE), which "reports the language that the respondent spoke at home, particularly if a language other than English was spoken."

The codebook provides the values for every LANGUAGED code:

```{r}
info.variables %>% 
  filter(var_name == 'LANGUAGE') %>% 
  pull(val_labels) %>% 
  `[[`(1)
```

## The data

All of the data in the extract. Most of these come preselected with any IPUMS extract.

```{r}
data %>% head()
```

The data is already filtered down to California. Filter it down again to just L.A. County, then keep only the relevant columns.

```{r}
la.data = data %>% 
  filter(COUNTYFIP == 37) %>% 
  select(YEAR, LANGUAGE, PERWT)

la.data %>% head()
```

## Aggregating microdata into totals

Each line of a microdata file represents a person's actual response to the survey. [PERWT](https://usa.ipums.org/usa-action/variables/PERWT#description_section) is the approxmiation of how many people the line of data represents. It needs to be aggregated and summed to get total counts for the language for that year.

```{r}
la.data.agg = la.data %>% 
  group_by(YEAR, LANGUAGE) %>% 
  summarise(PERWT = sum(PERWT))

la.data.agg %>% head()
```

This simple grouping and summing does **not** account for the margins of error associated with the estimates, which are important to consider. For those, I referred to [the result of the equivalent query](data/query-la-county-all-languages.pdf) on the IPUMS online SDA querying tool. The results of this microdata analysis and the SDA query are nearly identical.

Next, reformat the data, separating the labels from the code, plus add columns for the percent of the population speaking that language in each year and the county rank for that year.

```{r}
la.languages = la.data.agg %>% 
  group_by(YEAR) %>% 
  mutate(percent = PERWT / sum(PERWT)) %>% 
  # take out "N/A or blank" before calculating rank in year
  mutate(
    rankinyear = rank(
      desc(if_else(LANGUAGE != 0, percent, NA_real_)),
      na.last = 'keep',
      ties.method = 'min'
    )
  ) %>% 
  ungroup(YEAR) %>% 
  transmute(
    year = YEAR,
    langcode = zap_labels(LANGUAGE),
    language = as.character(as_factor(LANGUAGE)),
    total = PERWT,
    percent, rankinyear
  )

la.languages %>% head()
```

A quick check to ground truth the data: If you sum up the totals by year, does it come out to the county population over time?

```{r}
la.languages %>% 
  group_by(year) %>% 
  summarise(total = sum(total))
```

This looks [about right](https://en.wikipedia.org/wiki/Demographics_of_Los_Angeles_County). Populations will differ slightly based on these totals coming from aggregating person-level responses, but we are close.

## Analysis

What were the top 10 languages spoken in 1980?

```{r}
la.languages %>% 
  filter(year == 1980) %>% 
  arrange(rankinyear) %>% 
  head(10)
```

What were the top 10 languages spoken in 2018?

```{r}
la.languages %>% 
  filter(year == 2018) %>% 
  arrange(rankinyear) %>% 
  head(10)
```

How have English and Spanish, far and away the top languages, changed over time? (A version of this chart ran with the final story.)

```{r}
plot.eng.esp.yearly = la.languages %>% 
  filter(language %in% c('English', 'Spanish')) %>% 
  ggplot(aes(year, percent * 100, color = language)) +
  geom_line() +
  geom_point() +
  xlab('Year') +
  ylab('Percent of county population') +
  ggtitle('English and Spanish speakers in L.A. County') +
  theme_minimal()

plot.eng.esp.yearly
```

How about the other languages?

Start by keeping any language that has been in the top 10 for a particular year.

```{r message=FALSE}
ever.top.10 = la.languages %>% 
  filter(rankinyear <= 10) %>% 
  distinct(language) %>% 
  left_join(la.languages)

ever.top.10 %>% head()
```

```{r}
unique(ever.top.10$language)
```

This leaves 13 languages to look at more closely.

Let's take out English and Spanish. How has the percentage of the population speaking a particular language changed over time?

```{r}
plot.lang.yearly.bars = ever.top.10 %>%
  # take out the two biggest
  filter(!language %in% c('English', 'Spanish')) %>% 
  arrange(-year, -total) %>% 
  # control ordering
  mutate(language = fct_inorder(language)) %>% 
  ggplot(aes(as_factor(year), percent * 100, fill = language)) +
  geom_bar(stat = 'identity', position = 'dodge') +
  scale_fill_brewer(palette = "Paired", name = 'Language') +
  xlab('Year') +
  ylab('Percent of county population') +
  ggtitle('All languages ever in the top 10 spoken in L.A. County') +
  theme_minimal()

plot.lang.yearly.bars
```

What has the top 10 looked like over time?

```{r warning=FALSE}
plot.ranks.by.year = ever.top.10 %>%
  mutate(language = fct_inorder(language)) %>% 
  ggplot(aes(x = year, y = rankinyear, group = language)) +
  geom_hline(yintercept = 10.5, linetype = 2, alpha = 0.3) +
  geom_point(aes(size = total), color = 'grey') +
  geom_line(color = 'grey') +
  geom_point(
    data = . %>% filter(rankinyear > 4),
    aes(color = language, size = total)
  ) +
  geom_line(
    data = . %>% filter(rankinyear > 4),
    aes(color = language)
  ) +
  scale_x_continuous(
    limits = c(1980, 2021),
    breaks = c(1980, 1990, 2000, 2010, 2018),
    minor_breaks = NULL
  ) +
  scale_colour_brewer(palette = 'Paired') +
  scale_y_reverse(
    breaks = 1:10,
    limits = c(10,1),
    minor_breaks = NULL
  ) +
  theme_minimal() +
  theme(legend.position = 'none') +
  ylab('Rank in year') +
  xlab('') + 
  ggtitle('Top 10 languages spoken in L.A. County')

plot.top.10.by.year = plot.ranks.by.year +
  # annotate with names at end
  geom_text(
    data = . %>% group_by(language) %>% filter(rankinyear <= 10) %>% filter(year == max(year)),
    aes(year + 0.76, rankinyear, label = str_replace_all(word(language), ',', '')),
    hjust = 'left',
    size = 3
  )

plot.top.10.by.year
```

How have these languages risen and fallen in and out of the top 10?

```{r message=FALSE}
plot.top.10.rise.fall = plot.ranks.by.year +
  scale_y_reverse(
    breaks = 1:25,
    minor_breaks = NULL
  ) +
  # annotation
  geom_text(
    data = . %>% group_by(language) %>% filter(year == max(year)),
    aes(year + 0.76, rankinyear, label = str_replace_all(word(language), ',', '')),
    hjust = 'left',
    size = 3
  ) +
  ggtitle('All languages that have recently been in the top 10 for L.A. County')

plot.top.10.rise.fall
```

Summarize the change in languages other than English and Spanish by calculating the difference between now and 1980. (A version of this chart ran with the final story.)

```{r}
plot.change.1980.2018 = ever.top.10 %>% 
  filter(!language %in% c('English', 'Spanish')) %>%
  filter(year == 1980 | year == 2018) %>% 
  arrange(-total) %>% 
  select(language, year, percent) %>% 
  mutate(
    percent = percent * 100,
    language = word(language),
    language = str_replace_all(language, ',', ''),
    language = fct_rev(fct_inorder(language))
  ) %>% 
  ggplot(aes(x = percent, y = language)) +
  geom_segment(
    data = . %>%
      pivot_wider(names_from = year, values_from = percent) %>% 
      mutate(netgain = (`2018` - `1980`) > 0),
    aes(x = `1980`, xend = `2018`, y = language, yend = language, color = netgain),
    arrow = arrow(length = unit(0.2, "cm"))
  ) +
  geom_text(
    data = . %>% filter(language == 'Chinese' & year == 1980),
    aes(percent - 0.15, language, label = '1980'),
    size = 3.5
  ) +
  geom_text(
    data = . %>% filter(language == 'Chinese' & year == 2018),
    aes(percent + 0.15, language, label = '2018'),
    size = 3.5
  ) +
  theme_minimal() +
  theme(legend.position = 'none') +
  xlab('Percent of county population') +
  ylab('Language') +
  ggtitle('Change in percentage of L.A. county population by language')

plot.change.1980.2018
```

Write out the cleaned file:

```{r}
la.languages %>% write_csv('los-angeles-county-languages.csv', na = '')
```

